Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-02 _tex/index.tex
Built site for gh-pages

Nevertheless, the structural approach offers an additional key advantage
over the potential outcomes approach: it enables the graphical
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
can effectively convey the assumed causal structure of a system. They do
not represent detailed statistical models but allow researchers to
deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAG is accurate
\citep{McElreath_2020}.

\subsection{Directed acyclic graphs (DAG)}\label{sec-framework-dag}

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a formal and
graphical representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates that the edges of the graph extend from
one node to another, with arrows showing the direction of causal
influence. Moreover, the term \emph{acyclic} indicates the causal
influences do not form a loop, meaning the influences do not cycle back
on themselves \citep{McElreath_2020}.

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020, Morgan_et_al_2014}.
Figure~\ref{fig-dags} illustrates these blocks. Figure~\ref{fig-dag_bb1}
depicts two unconnected nodes, representing an scenario where variables
\(X_{1}\) and \(X_{2}\) are not causally related.
Figure~\ref{fig-dag_bb2} shows two connected nodes, illustrating a
scenario where a parent node \(X_{1}\) exerts a causal influence on a
child node \(X_{2}\). Consequently, \(X_{2}\) is considered a
\emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3} depicts a
\emph{chain} (or \emph{pipe}), where \(X_{1}\) influences \(X_{2}\), and
\(X_{2}\) influences \(X_{3}\). In this configuration, \(X_{1}\) is a
parent node of \(X_{2}\), and \(X_{2}\) a parent node of \(X_{3}\).
Furthermore, the DAG shows that \(X_{1}\) is an \emph{ancestor} of
\(X_{3}\) and that the relationship between these variables is entirely
\emph{mediated} by \(X_{2}\). Figure~\ref{fig-dag_bb4} illustrates a
\emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are both influenced
by \(X_{2}\). In this scenario \(X_{2}\) is a parent node of both
\(X_{1}\) and \(X_{3}\). Finally, Figure~\ref{fig-dag_bb5} depicts a
\emph{collider}, also known as \emph{inmorality}, where variables
\(X_{1}\) and \(X_{3}\) are concurrent causes of \(X_{2}\). In this
configuration, \(X_{1}\) and \(X_{3}\) are not causally related to each
other but both influence \(X_{2}\).

@@

\begin{minipage}{0.50\linewidth}

@@

\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb1.png}

@@

\subcaption{\label{fig-dag_bb1}Two unconnected nodes}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

@@

\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb2.png}

@@

\subcaption{\label{fig-dag_bb2}Two connected nodes or descendant}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb3.png}

@@

\subcaption{\label{fig-dag_bb3}Chain or pipe}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb4.png}

@@

\subcaption{\label{fig-dag_bb4}Fork}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb5.png}

@@

\subcaption{\label{fig-dag_bb5}Collider or inmorality}

@@

\caption{\label{fig-dags}DAG's fundamental building blocks.}

@@

@@

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
suggests that a variable \(X\) potentially influences both \(T\) and
\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
represent different treatments that could affect the recovery from
cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
the context of a DCJ study like the one described by
\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
child's cochlear implant use, which might influence the child's overall
speech quality \(Y\), with \(X\) indicating the child's hearing status.
Note that these descriptions do not specify the type of data or the
functional forms of the variables, highlighting the heuristic nature of
DAGs. {(not a bad example, but I prefer one using writing skills)}

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
graph, the variables of interest or \emph{endogenous} variables
\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
they are observed variables. The arrows in the graph reflect the
expected direction of causal influences among the variables. In
contrast, Figure~\ref{fig-dag_example2} depicts the \emph{magnified}
version of the DAG. This graph includes, in addition to the endogenous
variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
These exogenous variables are commonly referred to as
\emph{disturbances} or \emph{errors}, because they represent factors not
explicitly modeled. They are depicted as open circles, indicating their
unobserved nature. These exogenous variables are often omitted for
simplicity, resulting in the simplified DAG in
Figure~\ref{fig-dag_example1}. However, including them in the graph can
be advantageous in certain scenarios, as they can help to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. These concepts are introduced in the next
section.

Furthermore, a detailed examination of both figures reveals the presence
of at least four of the five fundamental building blocks.
Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept is introduced in the next section.

@@

\includegraphics[width=0.45\textwidth,height=\textheight]{images/figures/dag_example1.png}

@@

\subcaption{\label{fig-dag_example1}Simplied DAG}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_example2.png}

@@

\subcaption{\label{fig-dag_example2}Magnified DAG}

>>>>

Nevertheless, the structural approach offers an additional key advantage
over the potential outcomes approach: it enables the graphical
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system. A
heuristic is a strategy that simplifies information to make decisions
more quickly, efficiently, and sometimes more accurately than complex
methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
statistical models but allow researchers to deduce which statistical
models can provide valid causal inferences, assuming the causal
structure depicted in the DAG is accurate \citep{McElreath_2020}.

\subsection{DAGs and SCMs}\label{sec-framework-dag}

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) and its associated
structural causal models (SCM) to provide a graphical and formal
representation of the causal structure of a system \citep{Neal_2020}. In
this context, a \emph{graph} denotes a collection of nodes connected by
edges, where nodes represent random variables. The term \emph{directed}
indicates that the edges of the graph extend from one node to another,
with arrows showing the direction of causal influence. Moreover, the
term \emph{acyclic} indicates the causal influences do not form a loop,
meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

One key advantage of DAGs is that they can represent various causal
structures using only five fundamental building blocks, regardless of
complexity. The left panels of Figure~\ref{fig-dags_scms} illustrates
the \emph{simplified} graphs of these block. In these graphs, the
variables of interest or \emph{endogenous} variables, \(V=\{X,Z,Y\}\),
are depicted as solid black circles, indicating that they are observed
variables. The arrows in the graph reflect the expected direction of
causal influences among the variables.

In contrast, the middle panels of Figure~\ref{fig-dags_scms} depicts the
\emph{magnified} graphs of these building blocks. These graph include,
in addition to the endogenous variables, the \emph{exogenous} variables
\(E=\{e_{X},e_{Z},e_{Y}\}\). These exogenous variables, commonly
referred to as \emph{disturbances} or \emph{errors}, represent factors
not explicitly modeled. Moreover, they are depicted as open circles,
indicating their unobserved nature. These exogenous variables are often
omitted for simplicity, resulting in the simplified DAGs shown in the
left panels. However, including them in the graph can be advantageous in
certain scenarios, as they can help to highlight potential issues
related to conditioning and confounding \citep{Cinelli_et_al_2020},
concepts introduced in the next section.

@@

\begin{minipage}{0.33\linewidth}

@@

\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb1.png}

@@

\subcaption{\label{fig-sdag_bb1}Two unconnected nodes}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb1.png}

@@

\subcaption{\label{fig-mdag_bb1}Two unconnected nodes}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb1}Two unconnected nodes}

@@

\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb2.png}

@@

\subcaption{\label{fig-sdag_bb2}Two connected nodes or descendant}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb2.png}

@@

\subcaption{\label{fig-mdag_bb2}Two connected nodes or descendant}

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(X,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

@@

\subcaption{\label{fig-scm_bb2}Two connected nodes or descendant}

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb3.png}

\subcaption{\label{fig-sdag_bb3}Chain or pipe}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb3.png}

\subcaption{\label{fig-mdag_bb3}Chain or pipe}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb3}Chain or pipe}

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb4.png}

\subcaption{\label{fig-sdag_bb4}Fork}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb4.png}

\subcaption{\label{fig-mdag_bb4}Fork}

\[
\begin{aligned}
  X & := f_{X}(Z,e_{X}) \\
  Z & := f_{Z}(e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb4}Fork}

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb5.png}

\subcaption{\label{fig-sdag_bb5}Collider or inmorality}

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb5.png}

\subcaption{\label{fig-mdag_bb5}Collider or inmorality}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,Y,e_{Z}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb5}Collider or inmorality}

@@

\caption{\label{fig-dags_scms}The five fundamental building blocks of
DAGs with corresponding SCMs. \textbf{Note:} left column shows
simplified DAGS, middle column show magnified DAGs, and right column
shows the corresponding SCMs.}

@@

A careful examination of the left and middle panels of
Figure~\ref{fig-dags_scms} reveal what do the building blocks assume.
Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1} depict two unconnected
nodes, representing an scenario where variables \(X\) and \(Y\) are not
causally related. Figures \ref{fig-sdag_bb2} and \ref{fig-mdag_bb2} show
two connected nodes, illustrating a scenario where a parent node \(X\)
exerts a causal influence on a child node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-sdag_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain} (or \emph{pipe}), where \(X\)
influences \(Z\), and \(Z\) influences \(Y\). In this configuration,
\(X\) is a parent node of \(Z\), and \(Z\) a parent node of \(Y\).
Furthermore, the DAG shows that \(X\) is an \emph{ancestor} of \(Y\) and
that the relationship between these variables is entirely
\emph{mediated} by \(Z\). Figures \ref{fig-sdag_bb4} and
\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
\(Y\) are both influenced by \(Z\). In this scenario \(Z\) is a parent
node of both \(X\) and \(Y\). Finally, figures \ref{fig-sdag_bb5} and
\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
related to each other but both influence \(Z\).

@@

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
suggests that a variable \(X\) potentially influences both \(T\) and
\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
represent different treatments that could affect the recovery from
cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
the context of a DCJ study like the one described by
\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
child's cochlear implant use, which might influence the child's overall
speech quality \(Y\), with \(X\) indicating the child's hearing status.
{(not a bad example, but I prefer one using writing skills)}

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-sdag_example1} shows the \emph{simplified} graph, while
Figure~\ref{fig-mdag_example1} depicts the \emph{magnified} graph of the
DAG. A detailed examination of both figures reveals the presence of at
least four of the five fundamental building blocks.
Figure~\ref{fig-sdag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-mdag_example1}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept is introduced in the next section.

@@

\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_example1.png}

@@

\subcaption{\label{fig-sdag_example1}Simplied DAG}

@@

\includegraphics[width=0.65\textwidth,height=\textheight]{images/figures/mdag_example1.png}

@@

\subcaption{\label{fig-mdag_example1}Magnified DAG}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,Y,e_{Z}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_example1}Collider or inmorality}
